---
title: "Chapter 6: Programming in R"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
bibliography:
  - "`r system('kpsewhich mjandrews.bib', intern=TRUE)`"
biblio-style: apalike   
header-includes:
  - \input{header.tex}
---

```{r, echo=F}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, warning = FALSE, message = FALSE, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)
```

# Introduction

In this chapter, we aim to provide an introduction to some of the major programming features of R.

* We'll begin with *functions* both because they can be very simple to use and because of the major role functions play in programming in R generally.
* We will then consider *conditionals*, which allow us to execute different blocks of code depending on whether certain conditions are true.
* We will then consider *iterations*, also known as *loops*.
* This will lead on to *functionals*, which can often take the place of traditional loops in R.

Note, however, that there is more to programming in R than what we cover here.

# Functions

Let's say we have a vector of probabilities
$$
p_1, p_2 \ldots p_n
$$
and we want to calculate
$$
\log_b\left(\frac{p_i}{1-p_i}\right)\quad\text{for $i \in 1 \ldots n$}.
$$

We can define a function to implement this as follows:
```{r}
log_odds <- function(p, b){
  log(p/(1-p), base = b)
}
```

# Function input arguments

Notice that in the function definition, we stated that it takes two input arguments `p` and `b`, and that the code in the body explicitly operates on `p` and `b`.
As we see in the following code, we may also explicitly indicate which variables are mapped to `p` and `b`.

```{r}
probs <- c(0.1, 0.25, 0.5, 0.75)
log_base = 2
log_odds(p=probs, b=log_base)
```
When using explicit assignment like this, the order of the arguments no longer matters. Thus, we can write the above code as follows. 
```{r}
log_odds(b=log_base, p=probs)
```

# Anonymous functions

Consider the following function:
```{r}
f <- function(x, g){
  g(sum(x))
}
```
Here, `f` takes an object `x` and a function `g` and calls `g(sum(x))` and returns the result.
We can pass in any existing R function we wish as the value of `g`, as in the following examples.
```{r}
x <- c(0.1, 1.1, 2.7)
f(x, log10)
```
However, we don't have to assign a name to our custom function and pass in that name.
We can instead just pass in the unnamed, or *anonymous*, function itself as in the following examples.
```{r}
f(x, function(x) x^2)
```
Anonymous functions are widely used in R. 


# Conditionals

Conditionals allows us to execute some code based on whether some condition is true or not. Consider the following simple example.
```{r}
library(tidyverse)

# Make a data frame
data_df <- tibble(x = rnorm(10),
                  y = rnorm(10))


write_data <- TRUE

if (write_data) {
  write_csv(data_df, 'tmp_data.csv')
}
```

Here, we write `data_df` to a `.csv` file if and only if `write_data` is true.


# *if ... else* statments 

Sometimes, we want to execute one code block if the condition is true and execute an alternative code block if it is false.
To do this, we use an *if ... else* statement, as in the following example.
```{r, echo=F}
write_csv(tibble(x = 17),
          'data_new.csv')
write_csv(tibble(x = 42),
          'data_old.csv')
```
```{r}
use_new_data <- TRUE

if (use_new_data){
  data_df <- read_csv('data_new.csv')
} else {
  data_df <- read_csv('data_old.csv')
}
```
```{r, echo=F}
# Just to be sure that the above 
# code worked.
stopifnot(data_df$x == 17) 
```
As we can see, if `use_new_data` is true, we read in the data from `data_new.csv`, and otherwise we read the data in from `data_old.csv`. 

```{r, echo=F}
write_data_foo <- function(i){
  write_csv(tibble(x=i), paste0('data_', i, '.csv'))
} 
for (i in seq(3)) {write_data_foo(i)}
```


# `for` loops

In R, there are two types of iterations or *loops*, which we informally refer to as *`for` loops* and *`while` loops*.

Let's say we had the following vector of 1000 elements to which we wished to apply the `relu` function.
```{r}
N <- 1000
x <- seq(-0.1, 1.1, length.out = N)

relu <- function(x){
  if (x < 0) { 0 } else { x }
}
```
We can create a for loop as follows.
```{r}
y <- numeric(N)
for (i in 1:N) {
  y[i] <- relu(x[i])
}
```

For each value of `i` from `1` to `N`, we execute `y[i] <- relu(x[i])`.


# `while` loops

Unlike for loops, which iterate through each value in a sequence of values, while loops continuously execute the code body while a condition remains true.

As a very simple example of a while loop, let's say we wish to find the largest value of $k$ such that $2^k \leq 10^6$. 
```{r}
k <- 0
while (2^(k+1) <= 10^6) {
  k <- k + 1
}
k
```


# `lapply` 

Functionals are functions that take a function as input and return a vector.
There are many functionals in the base R language.
Here, we will look  at one of the most widely used: `lapply`

The `lapply` functional takes a vector and a function, applies the function to each element in the vector and returns a new list.
```{r}
x <- c(-1, 0, 1)
y <- lapply(x, relu)
y
```


# Functionals with `purrr`

The `purrr` package in the `tidyverse` provides functionals like `lapply`.
We can load `purrr` with `library(purrr)`, but it is also loaded by `library(tidyverse)`.

One of the main tools in `purrr` is `map` and its variants.
It is very similar to `lapply`.
```{r}
y <- map(x, relu)
y
```

