---
title: "Chapter 3: Data Wrangling"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
bibliography:
  - "`r system('kpsewhich mjandrews.bib', intern=TRUE)`"
biblio-style: apalike   
header-includes:
  - \input{header.tex}
---


```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, warning = FALSE, message = FALSE, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)

options(scipen=9)
```


```{r, echo=F}
library(tidyverse)
```

# Data wrangling with the `tidyverse`

* There are many tools in R for doing data wrangling. 
* Here, we will focus of a core set of inter-related `tidyverse` tools. 
* These include the commands available in the `dplyr` package, particularly its so-called *verbs* such as the following:
  * select
  * rename
  * slice
  * filter
  * mutate
  * arrange
  * group_by
  * summarize

# Data wrangling with the `tidyverse`

* In addition, `dplyr` provides tools for merging and joining data sets. 
* Next, there are the tools in the `tidyr` package, particularly the following:
  * gather
  * spread
  * unite
  * separate
  * pivot_longer
  * pivot_wider
* Other packages such as `lubridate` and `stringr` provide essential tools for dealing with and manipulating dates and strings, respectively.
* All of these tools are can be combined together using the `%>%` pipe operator

# The `dplyr` verbs

As an example data set, we will use the data contained in the file `blp-trials-short.txt`. 

```{r}
blp <- read_csv('data/blp-trials-short.txt')
```

# Select variables with `select`

The `dplyr` command `select` allows us to select columns from a data frame. 
For example, if we just want `participant`, `lex`, `resp`, and `rt`, then we would do the following.
```{r}
select(blp, participant, lex, resp, rt)
```

# Select variables with `select`

We can select a range of variables by specifying the first and last variables in the range with a `:` between them as follows.
```{r}
select(blp, spell:prev.rt)
```

# Select variables with `select`

We can also select a range of variables using indices as in the following example.
```{r}
select(blp, 2:5) # columns 2 to 5
```

# Select variables with `select`

We can select variables according to the character or characters that they begin with. For example, we select all variables that being with `p` as follows. 
```{r}
select(blp, starts_with('p'))
```

# Select variables with `select`

Or we can select variables by the characters they end with.
```{r}
select(blp, ends_with('t'))
```

# Select variables with `select`

We can select variables that contain a certain set of characters in any position. 
For example, the following selects variables whose names contain the string `rt`.
```{r}
select(blp, contains('rt'))
```


# Select variables with `select`

We can also match by regular expressions. For example, the regular expression `^rt|rt$` will match the `rt` if it begins or ends a string. Therefore, we can select the variables that contain `rt`, where the string `rt` means reaction time, as follows.
```{r}
select(blp, matches('^rt|rt$'))
```

# Remove variables with `select`

We can use `select` to *remove* variables as well as select them. To remove a variable, we precede its name with a minus sign. 
```{r}
select(blp, -participant) # remove `participant`
```

# Remove variables with `select`

Just as we selected ranges or sets of variables above, we can remove them by preceding their selection functions with minus signs.
```{r}
select(blp, -(2:6))
```

# Remove variables with `select`

Or, as another example, we can remove the variables that contain the string `rt` as follows.
```{r}
select(blp, -contains('rt'))
```

# Renaming variables with `select`

When we select individual variables with `select`, we can rename them too, as in the following example.
```{r}
select(blp, subject=participant, reaction_time=rt)
```

# Renaming variables with `rename`

If we want to rename some variables, and get a data frame with all variables, including the renamed ones, we should use `rename`.
```{r}
rename(blp, subject=participant, reaction_time=rt)
```


# Selecting observations by indices with `slice` 

We use `slice` to select observations by their indices. For example, 
to select rows 10, 20, 50, 100, 500, we would simply do the following.
```{r}
slice(blp, c(10, 20, 50, 100, 500))
```

# Selecting observations by indices with `slice` 

Given that, for example, `10:100` would list the integers 10 to 100 inclusive, we can select just these observations as follows.
```{r}
slice(blp, 10:100)
```

# De-selecting observations by indices with `slice` 

Just as we did with `select`, we can precede the indices with a minus sign to drop the corresponding observations. Thus, for example, we can drop the first 10 observations as follows.
```{r}
slice(blp, -(1:10))
```

# Selecting observations by condition with `filter`

The `filter` command is a powerful means to filter observations according to their values. For example, 
we can select all the observations where the `lex` variable is `N` as follows.
```{r}
filter(blp, lex == 'N')
```

# Selecting observations by condition with `filter`

We can also filter by multiple conditions by listing each one with commas between them. For example, the following gives us the observations where `lex` has the value of `N` and `resp` has the value of `W`.
```{r}
filter(blp, lex == 'N', resp=='W')
```


# Selecting observations by condition with `filter`

The following gives us those observations where where `lex` has the value of `N` and `resp` has the value of `W` and `rt.raw` is less than or equal to 500.
```{r}
filter(blp, lex == 'N', resp=='W', rt.raw <= 500)
```

# Selecting observations by condition with `filter`

The previous command is equivalent to making a conjunction of conditions using `&` as follows.
```{r}
filter(blp, lex == 'N' & resp=='W' & rt.raw <= 500)
```

# Selecting observations by condition with `filter`

We can make a *disjunction* of conditions for filtering using the logical-or symbol `|`. For example, 
to filter observation where the `rt.raw` was either less than 500 or greater than 1000, we can
do the following.
```{r}
filter(blp, rt.raw < 500 | rt.raw > 1000)
```


# Changing variables and values with `mutate`

To create a new variable `is_accurate` that takes the value of `TRUE`
whenever `lex` and `resp` have the same value, we can do the following:
```{r}
mutate(blp, acc = lex == resp)
```

# Changing variables and values with `mutate`

As another example, we can create a new variable that gives the length of the word given by the `spell` variable.
```{r}
mutate(blp, len = str_length(spell))
```


# Changing variables and values with `mutate`

We can also create multiple new variable at the same time as in the following example.
```{r}
mutate(blp, 
       acc = lex == resp, 
       fast = rt.raw < mean(rt.raw, na.rm=TRUE))
```



# Sorting observations with `arrange`

Sorting observations in a data frame is easily accomplished with `arrange`. For
example to sort by `participant` and then by `spell`, we would do the following.
```{r}
arrange(blp, participant, spell)
```

# Sorting observations with `arrange`

We can sort by the reverse order of any variable by using the `desc` command on the variable. In the following example,
we sort by `participant`, and then by `spell` in reverse order.
```{r}
arrange(blp, participant, desc(spell))
```


# Reducing data with `summary`

The `dplyr` package has a function `summarize` (or, equivalently, `summarise`)
that applies summarizing functions to variables. 

For example, we may calculate some summary statistics of the particular variables as in the following example.
```{r}
summarize(blp, 
          mean_rt = mean(rt, na.rm = T),
          median_rt = median(rt, na.rm = T),
          sd_rt.raw = sd(rt.raw, na.rm = T)
)
```
(Note that here it is necessary to use `na.rm = T` to remove the `NA` values in the variables.)

# Reducing data with `summary` and `group_by`

The `summarize` command, and its variants, become considerably more powerful
when combined with the `group_by` command. Effectively, `group_by` groups the
observations within a data frame according to the values of specified variables.
For example, the following command groups `blp` into groups of observations
according to value of the `lex` variable.
```{r}
blp_by_lex <- group_by(blp, lex)
```

# Reducing data with `summary` and `group_by`

If we now apply `summarize` to this grouped data frame, we will obtain summary
statistics for each group, as in the following example.
```{r}
summarize(blp_by_lex, mean = mean(rt, na.rm=T))
```


# The %>% operator

The `%>%` operator in R is known as the *pipe*. It available from the `magrittr` package, which is part of the `tidyverse`. In RStudio, the keyboard shortcut Ctrl+Shift+M types `%>%`.

To understand pipes, let us begin with a very simple example. The following
`primes` variable is a vector of the first 10 prime numbers.
```{r}
primes <- c(2, 3, 5, 7, 11, 13, 17, 19, 23, 29)
```
We can calculate the sum of `primes` as follows.
```{r}
sum(primes)
```
We may then calculate the square root of this sum.
```{r}
sqrt(sum(primes))
```
We may then calculate the logarithm of this square root.
```{r}
log(sqrt(sum(primes)))
```


#  The %>% operator

The `%>%` is *syntactic sugar* that reexpresses nested functions as sequences. 

Returning to some of our examples above, we will see how they can be rewritten with pipes. In each case,
we will precede the piped version with a comment showing its original version.
```{r}
# sum(primes)
primes %>% sum()
```
```{r}
# sum(primes, na.rm=T)
primes %>% sum(na.rm=T)
```
```{r}
# log(sqrt(sum(primes)))
primes %>% sum() %>% sqrt() %>% log()
```



# Reshaping with `pivot_longer` and `pivot_wider`

A so-called *tidy* data set is a data set where all rows are observations, all columns are variables, and each variable is a single value.

Consider the following data frame.
```{r}
recall_df <- read_csv('data/repeated_measured_a.csv')
recall_df
``` 
In this data frame, for each subject, we have three values, which are their scores on a memory test in three different conditions of an experiment.
The conditions are `Neg` (negative), `Neu` (neutral), `Pos` (positive).
However, each column is not a variable. 
The `Neg`, `Neu`, `Pos` are, in fact, *values* of a variable, namely the condition of the experiment.

# Reshaping with `pivot_longer` 

To tidy this data frame, we need a variable for the subject, another for the experiment's condition, and another for the memory score for the corresponding subject in the corresponding condition.
To do so, we perform what is sometimes known as a *wide to long*  transformation.
The `tidyr` package has a function `pivot_longer` for this transformation.

To use `pivot_longer`, we must specify the variables (using the `cols` argument) that we want to pivot from wide to long.
Next, we must provide a name for the column that will indicate the experimental condition.
Finally, we must provide a name for the column that will indicate the memory scores.

# Reshaping with `pivot_longer` 

Here is the necessary code:
```{r}
recall_long <- pivot_longer(
  recall_df, cols = -Subject, names_to = 'condition', 
  values_to = 'score')
recall_long
```

# Reshaping with `pivot_wider` 

The inverse of a `pivot_longer` is a `pivot_wider`.
It is very similar to `pivot_longer` and we use `names_from` and `values_from` in the opposite sense to `names_to` and `values_to`.
```{r}
pivot_wider(recall_long, 
            names_from = 'condition', values_from = 'score')
```


