---
title: "Data Analysis: Report II"
author: "Mark Andrews"
date: "October 25, 2019"
output: 
  pdf_document:
    keep_tex: yes
header-includes:
- \usepackage{booktabs}
bibliography: refs.bib
biblio-style: apalike  
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r load_packages, echo = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

weight_df <- read_csv('weight.csv')
weight_df_grouped <- group_by(weight_df, gender)

sample_size <- weight_df_grouped %>%
  summarise(n=n()) %>% 
  deframe()

weight_df_summary <- weight_df_grouped %>% 
  summarise_at(vars(weight, height), 
               list(avg=mean, stdev=sd))
```

# Descriptive statistics

In this data set, we have measured the weight (in kg)
and height (in cm) of `r sum(sample_size)` participants 
(`r sample_size['Male']` males, `r sample_size['Female']`
females). In the following table, we show the mean and
the standard deviations of the weights and heights, 
separately for males and females. 
                          
```{r descriptives_table, echo=F}
weight_df_summary %>% 
  kable(format = "latex", 
        booktabs = TRUE, 
        digits = 2, 
        align = 'c') %>%
    kable_styling(position = "center")
```

# Statistical model

We will model the relationship between weight and height 
as a varying intercepts normal linear model as follows. 
For each $i \in 1 \ldots n$, 
$$
\begin{aligned}
y_i &\sim N(\mu_i, \sigma^2),\\
\mu_i &= \beta_0 + \beta_1 x_i + \beta_2 z_i, 
\end{aligned}
$$
where $y_i$, $x_i$, $z_i$ are the weight, height, 
and gender of participant $i$.

In R, this analysis can be easily performed as follows.
```{r stat_model}
model <- lm(weight ~ height + gender, data = weight_df)
```
```{r model_results, echo=F}
R_sq <- summary(model)$r.sq
f_stat <- summary(model)$fstatistic
p_value <- pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = F)
```

The $R^2$ for this model is `r round(R_sq, 2)`,
$F(`r round(f_stat[2])`, `r round(f_stat[3])`) = 
`r round(f_stat[1],2)`$,
$p `r format.pval(p_value, eps = 0.01)`$.


More information about varying intecept models can be
found in @GelmanHill:2007.

# References
