---
title: "Chapter 5: Exploratory Data Analysis"
author: "Mark Andrews"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: true
    highlight: haddock
  html_document:
    highlight: haddock
bibliography:
  - "`r system('kpsewhich mjandrews.bib', intern=TRUE)`"
biblio-style: apalike   
header-includes:
  - \input{header.tex}
---

  
  
```{r, echo=F}
knitr::opts_chunk$set(echo = F, prompt = F, warning = F, message = F, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)

# for cropped pdf files
system("mkdir -p include_graphics")
```

```{r}
library(tidyverse)
library(cowplot)
theme_set(theme_classic())
```

# Introduction

In his famous 1977 book *Exploratory Data Analysis*, John Tukey describes exploratory data analysis as detective work.
He likens the data analyst to police investigators who look for and meticulously collect vital evidence from a crime scene. 
By contrast, he describes the task undertaken within the courts system of making a prosecution case and evaluating evidence for or against the case as analogous to confirmatory data analysis. 
In confirmatory data analysis, we propose models of the data and the evaluate these models.
This is almost always the ultimate goal of data analysis, but just as good detective work is necessary for the successful operation of the judicial system, exploratory data analysis is a vital first step prior to confirmatory data analysis and data modelling generally. 
As Tukey puts it "It is important to understand what you can do before you learn to measure how well you seem to have done it." (@tukey1977exploratory; Page v).
Similar perspectives are to be found in @hartwig1979exploratory who argue that the more we understand our data, the "more effectively data can be used to develop, test, and refine theory." (@hartwig1979exploratory; Page 9).

In this book, we are wholly committed to the idea that probabilistic modelling is the ultimate goal and even the raison d'être of data analysis. 
Part II of this book is devoted to covering this topic in depth.
However, and in line with the advice of @tukey1977exploratory, @hartwig1979exploratory, and others, this probabilistic modelling can only be done thoroughly and well if we understand our data.
In particular, understanding the data allows us to select suitable and customized models for the data, rather than naively choosing familiar off-the-shelf models that may be either limited or inappropriate.

<!--  exploring univariate and bivariate data. -->
<!-- Although almost all cases of real world data sets contain more than just univariate or bivariate data, as we explain, understanding univariate and bivariate data serve as the foundation for understanding much more complex data sets. -->

# Univariate data

In this chapter, we describe quantitative and graphical methods for exploratory data analysis by focusing on *univariate data*.
Univariate data is data concerning a single variable.
For example, if we record the estimated market value of each house in a sample of houses and record nothing other than this value, then our dataset would be a univariate one.
Put like this, it is rare indeed for real world data sets to be truly univariate.
In any data set, however small, we almost always have values concerning more than one variable.
Nonetheless, understanding univariate data has a fundamental role in data analysis.
First of all, even when we do have data concerning dozens or even hundreds of variables, we often need to understand individual variables in isolation.
In other words, we often first explore or analyse our many variables individually before examining the possible relationships between them.
More importantly, however, is that often our ultimate goal when analysing data is to perform *conditionally univariate* analyses.
In other words, we are often interested in understanding how the distribution of one variable changes given, or conditioned upon, the values of one or many other variables.
For example, to use our house price example, we may be ultimately interested in understanding house prices, and all the other variables that we collect --- such as when the house was first built, how many bedrooms it has, where it is located, whether it is detached or semi-detached or terraced, and so on --- are all used to understand how the distribution of prices vary as a function of the age of the house, the number of bedrooms, the neighbourhood, and so on.
Conditionally univariate analysis of this kind is, in fact, what is being done in all regression analyses that considers a single outcome variable at a time.
This encompass virtually all the well known type of regression analyses, whether based on linear, or generalized linear, or multilevel models.
Almost all of the many examples of regression analyses that we consider in Part II of this book are examples of this kind.
Conditionally univariate analysis is also a major part of exploratory data analysis: we are often interested in exploring, whether by visualizations or using summary statistics, how one variable is distributed when other variables are held constant.
In fact, many cases of what we might term bivariate or multivariate exploratory analysis may be more accurately described as conditionally univariate exploratory analyses.

## Types of univariate data

In a famous paper, @stevens:measurement defined four (univariate) data types that are characterized by whether their so-called *levels of measurement* is either *nominal*, *ordinal*, *interval*, and *ratio*.
Roughly speaking, according to this account, nominal data consist of values that are names or labels.
Ordinal data consist of values that has some natural ordering or rank.
The interval and ratio data types are types of continuous variables.
Ratio are distinguished from interval scales by whether they have a non-arbitrary, or true, zero or not, respectively.
Although this typology is still seen as a definitive in psychology and some other social science fields, it is not one that we will use here because is not widely endorsed, or even widely known, in statistics and data analysis generally.
In fact, almost from its inception, it has been criticized as overly strict, limited, and likely to lead to poor data analyses practice [see, for example, @velleman:measurement, and references therein].
In general in statistics or data science, there is no one definitive taxonomy of data types.
Different, though often related, categorization abound, with each one having finer or coarser data types distinctions.
Nonetheless, it is worthwhile to make these typological distinctions because different types of data are often best described, visualized, and eventually statistically modelled using different approaches and techniques.
For example, when and why to use the different types of regression analyses described in Part II of this book is often based solely on what type of data we are trying to model.
The following data types are based on distinctions that are very widely or almost universally held.
In each case, however, their definitions are rather informal, and whether any given data is best classified as one type or another is not always clear or uncontroversial.

Continuous data

~ *Continuous data* represents the values or observations of variable that can take any value in a continuous metric space such as the real line or some interval thereof.
Informally speaking, we'll say a variable is continuous if its values can be ordered and between any two values, there exists another value and hence an infinite number of other values.
A person's height, weight, or even age are usually taken to be continuous variables, as are variables like speed, time, distance.
All these examples take on only positive values, but others like a bank account's balance, or temperature (on Celsius or Fahrenheit scale, for example) can also take on negative values.
Another important subtype of the continuous variables are proportions, percentages, or probabilities.
These exist strictly on the interval of the real line between $0$ and $1$.
In their data typology, @mosteller1977data consider these as part of a special data type in itself, but here, we'll just treat them as a subtype of the continuous data variables.

Categorical data 

~ *Categorical data* is where each value takes on one of a (usually, but not necessarily) finite number of values that are categorically distinct and so are not ordered, nor do they exist as points on some interval or metric space.
Examples include a person's nationality, country of residence, occupation.
Likewise, subjects or participants in an experiments or study are assigned categorically distinct identifiers, and could be assigned to categorically distinct experimental conditions, such a control group versus treatment group, etc.
Values of categorical variables can not naturally or uncontroversially be placed in a order, nor is there is any natural or uncontroversial sense of distance between them.
The values of a categorical variable are usually names or labels, and hence *nominal data* is another widely used term for categorical data.
A categorical variable that can only take one of two possible values --- control versus treatment, correct versus incorrect, etc --- is usually referred to simply as binary or dichotomous variables.
Categorical variables taking on more than two variables are sometimes called polychotomous.


Ordinal data 

~ *Ordinal data* represent values of a variable that can be ordered but have no natural or uncontroversial sense of distance between them.
First, second, third, and so on, are values with a natural order, but there is no general sense of distance between them.
For example, knowing that three students scored in first, second, and third place, respectively, on an exam tells us nothing about how far apart their scores were.
All we generally know is that the student who came first had a higher score than the student who scored second, who had a higher score the student who came third.
Ordinal variable may be represented by numbers, for example, first, second, and third place could be represented by the numbers 1, 2, and 3.
These numbers have their natural order, i.e. $1 < 2 < 3$, but they do not exist a metric space.

Count data

~ *Count data* are tallies of the number of times something has happened or some value of a variable has occurred.
The number of cars stopped at a traffic light, the number of people in a building, the number of questions answered correctly on an exam, and so on, are all counts.
In each case, the values take on non negative integer values.
In general, they have a lower bound of zero, but do not necessarily have an upper bound.
In some cases, like the number of questions answered correctly on an exam with 100 questions, there is also an upper bound.
In other cases, like the number of shark attacks at Bondi beach in any given year, does not have a defined upper value.
The values of a count variable are obviously ordered, but also have a true sense of distance.
For example, a score of 87 correct answers on an exam is as far from a score of 90 as a score of 97 is from 100, and so on.

## Characterizing univariate distributions 


We can describe any univariate distribution in terms of three major features: *location*, *spread*, and *shape*.
We will explore each of these in detail below through examples, but they can be defined roughly as follows.

Location

~ The *location* or *central tendency* of a distribution describes, in general, where the mass of the distribution is located on an interval or along a range of possible values.
More specifically, it describes the typical or central values that characterize the distribution.
Adding a constant to all values of a distribution will change the location of the distribution, by essentially shifting the distribution rigidly to left or to the right.
  
Dispersion 

~ The *dispersion*, *scale*, or *spread* of a distribution of a distribution tells us how dispersed or spread out the distribution is.
It tells us roughly how much variation there is in the distribution's values, or how far apart are the values from one another on average.
   
Shape

~ The shape of a distribution is roughly anything that is described by neither the location or spread.
Two of the most important shape characteristics are *skewness* and *kurtosis*.
Skewness tells us how much asymmetry there is in the distribution.
A left or negative skew means that the tail on the left (that which points in the negative direction is) is longer than that on the right, and this entails that the center of mass is more to the right and in the positive direction.
Right or positive skew is defined by a long tail to the right, or in the positive direction, and hence the distribution is more massed to the left.
Kurtosis is a measure of how much mass is in the center versus the tails of the distribution.

# Measures of central tendency

Let us assume that we have a sample of $n$ univariate values denoted by $x_1, x_2 \ldots x_i \ldots x_n$. 
Three commonly used measures of central tendency, at least in introductory approaches to statistics, are the arithmetic mean, the median, and the mode. 
Let us examine each in turn.

## Arithmetic Mean

The arithmetic mean, or usually known as simply *the* mean, is defined as follows:
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i.
$$
It can be seen as the centre of gravity of the set of values in the sample.
This is not simply a metaphor to guide our intuition. 
For example, if we had $n$ point masses, each of equal mass, positioned at points $\omega_1, \omega_2 \ldots \omega_n$ along some linear continuum, then $\Omega = \frac{1}{n} \sum_{i=1}^n \omega_i$ is their centre of gravity. 




The arithmetic mean is also the finite sample counterpart of the *mathematical expectation* or *expected value* of a random variable. 
If $X$ is a continuous random variable with probability distribution $\Prob{X}$, then the expected value of $X$ is defined as 
$$
\langle X \rangle = \int_{-\infty}^\infty x \Prob{X = x} dx.
$$
On the other hand, if $X$ takes on $K$ discrete values, then its expected value is
$$
\langle X \rangle = \sum_{k=1}^K x_k \Prob{X = x_k}.
$$
A finite sample of points $x_1, x_2 \ldots x_i \ldots x_n$ can be represented as a discrete probability distribution of a random variable $X$ with a finite number of values such that $\Prob{X = x_i} = \frac{1}{n}$.
Therefore,
$$
\langle X \rangle = \sum^n_{i=1} x_i \Prob{X = x_i} = \frac{1}{n} \sum^n_{i=1} x_i  = \bar{x}.
$$


As we will discuss in more detail below, the mean is highly sensitive to outliers, which we will define for now as simply highly atypical values.
As an example, consider the following human reaction time (in milliseconds) data.
```{r, echo=T}
rt_data <- c(567, 1823,  517,  583,  317,  367,  250,  503,
             317,  567,  583,  517,  650,  567,  450,  350)
```
The mean of this data is 
```{r, echo=T}
mean(rt_data)
```
This value is depicted by the red dot in Figure \ref{fig:rt_mean}.
However, here there is an undue influence of the relatively high second value.
If we remove this value, the mean becomes much lower.
```{r,echo=T}
rt_data[-2] %>% mean()
```
This new mean is shown by the blue dot in Figure \ref{fig:rt_mean}.

```{r}
set.seed(47)
p1 <- ggplot(enframe(rt_data),
       aes(x = '', y = value)
) + #geom_boxplot(width=0.33) +
  #scale_y_continuous(limits=c(.25, 1)) + 
  geom_jitter(size=0.5, width = 0.2, alpha = 0.5) + 
  scale_x_discrete(breaks=NULL) + 
  coord_flip() +
  theme_classic() +
  theme(aspect.ratio = 1/5) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  geom_point(data = data.frame(x = 1.0, y=mean(rt_data)),
             aes(x = x, y = y), col='red', size=1.5) +
  geom_point(data = data.frame(x = 1.0, y=mean(rt_data[-2])),
             aes(x = x, y = y), col='royalblue', size=1.5) +
  ylab('Reaction time')

ggsave("rt_mean.pdf", p1, device = 'pdf')
system("./crop_pdf  rt_mean.pdf include_graphics/rt_mean-crop.pdf")
system("rm rt_mean.pdf")
```

```{r rt_mean, fig.cap='The red dot shows the mean value of the set of points in black. In this case, there is an undue influence of the one relatively large value. The blue dot displays the mean of the sample after that one large value is removed.', fig.align='center'}
knitr::include_graphics('include_graphics/rt_mean-crop.pdf')
```

## Median

The median of a finite sample is defined as the middle point in the sorted list of its values.
If there is an odd number of values, there is exactly one point in the middle of the sorted list of values.
If there is an even number of values, there there are two points in the middle of the of the sorted list.
In this case, the median is the arithmetic mean of these two points.
In the `rt_data` data set, the median is as follows.
```{r, echo=T}
median(rt_data)
```

The sample median just defined is the finite sample counterpart of the median of a random variable. 
Again, if $X$ is a random variable with probability distribution $\Prob{X}$, then its median is defined as the value $m$ that satisfies 
$$
\int_{-\infty}^m \Prob{X = x} dx = \frac{1}{2}.
$$


```{r}
set.seed(47)
p1 <- ggplot(enframe(rt_data),
       aes(x = '', y = value)
) + #geom_boxplot(width=0.33) +
  #scale_y_continuous(limits=c(.25, 1)) + 
  geom_jitter(size=0.5, width = 0.2, alpha = 0.5) + 
  scale_x_discrete(breaks=NULL) + 
  coord_flip() +
  theme_classic() +
  theme(aspect.ratio = 1/5) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  geom_point(data = data.frame(x = 1.0, y=median(rt_data)),
             aes(x = x, y = y), col='red', size=1.5) +
  ylab('Reaction time')

ggsave("rt_median.pdf", p1, device = 'pdf')

system("./crop_pdf rt_median.pdf include_graphics/rt_median-crop.pdf")

system("rm rt_median.pdf")
```

```{r rt_median, fig.cap='The red dot shows the median value of the set of points in black. In this case, unlike with the mean, there is no undue influence of the one relatively large value. ', fig.align='center'}
knitr::include_graphics('include_graphics/rt_median-crop.pdf')
```

Unlike the mean, the median is robust to outliers. 
In Figure \ref{fig:rt_median}, we see how the median is not unduly influenced by the presence of the one extreme value.
In fact, if this point were removed, the median would be unaffected.

## Mode

The sample mode is the value with the highest frequency. 
When dealing with random variables, the mode is clearly defined as the value that has the highest probability mass or density.
For example, if $X$ is a continuous variable, then the mode is given as follows:
$$
\textrm{mode} = \argmax_{x} \Prob{X = x},
$$
which is the value of $x$ for which the density $\Prob{X=x}$ is at its maximum.
For a discrete variable, the mode is as similarly defined.
$$
\textrm{mode} = \argmax_{x_k} \Prob{X = x_k} dx,
$$

While the mode is clearly defined for random variables, for finite samples, it is in fact not a simply matter.
For example, using our `rt_data` data above, we can calculate the frequency of occurrence of all values as follows.
```{r, echo=T}
table(rt_data)
```
Clearly, in this case, most values occur exactly once.
We can identify the value that occurs most often as follow.
```{r, echo=T}
which.max(table(rt_data)) %>% names()
```

In practice, it often occurs that all values in the data set occur exactly once.
This is especially the case when values can be anything in a wide range of values.
In these cases, there is essentially no mode, or the mode can be said to be undefined.
One strategy to deal with these situations is to bin the data first. 
To illustrate this, let us sample $10000$ values from a normal distribution with a mean of $100$ and standard deviation of $50$.
```{r, echo=T}
N <- 10000
x <- rnorm(N, mean = 100, sd = 50)
```
We know that the mode of this normal distribution is $100$.
However, because the values generated by `rnorm` are specified with many decimal digits, we are astronomically unlikely to get any more than 1 occurrence per each unique value.
We can confirm this easily.
```{r, echo=T}
length(unique(x)) == N
```
One possibility in this case would be to bin these values into, say, 50 equal sized bins, and then find the mid value of the bin with the highest frequency.
We may do this using the `hist` function. 
This is usually used for plotting, but also can return counts of values that are binned, and we can suppress the plot with `plot = F`.
```{r, echo=T}
H <- hist(x, 50, plot=F)
```
We can use `which.max` applied to the `counts` property of the object that `hist` returns.
This will give us the index of the bin with the highest frequency, and then we can use the `mids` property to get the mid point of that bin.
We can put this is a function as follows.
```{r, echo=T}
sample_mode <- function(x, n_bins=10){
  h <- hist(x, breaks = n_bins, plot = F)
  h$mids[which.max(h$counts)]
}

sample_mode(x)
sample_mode(x, n_bins = 50)
sample_mode(x, n_bins = 100)
```

## Robust measures of central tendency

We've seen that the mean is unduly influenced by outliers.
More technically, the mean has a low *breakdown* point.
The breakdown point of a statistic is the proportion of values in a sample that be arbitrarily large before the statistic becomes arbitrarily large.
The mean has a breakdown point of zero.
By making just one value arbitrarily large, the mean becomes arbitrarily large.
By contrast, the median has a very high breakdown point.
In fact up to 50% of the values of sample can be arbitrarily large before the median becomes arbitrarily large.
The median, therefore, is said to be a very robust statistic, and the mean is a particularly unrobust statistic.

There are versions of the standard arithmetic mean that are specifically designed to be robust.
The *trimmed mean* removes a certain percentage of values from each extreme of the distribution before calculating the mean as normal.
The following code, for example, removes 10% of values from the high and low extremes of `rt_data`.
```{r, echo=T}
mean(rt_data, trim = 0.10)
```
The trimmed mean can be calculated using the following function.
```{r, echo=T}
trimmed_mean <- function(x, trim = 0.1){
  
  n <- length(x)
  lo <- floor(n * trim) + 1
  hi <- n + 1 - lo
  sort(x)[lo:hi] %>% 
    mean()
  
}
```
Thus, we see that in the case of `rt_data`, where $n = `r length(rt_data)`$, that the trimmed mean is based on elements `r floor(length(rt_data) * 0.1) + 1` to `r length(rt_data) + 1 - (floor(length(rt_data) * 0.1) + 1)` of `rt_data` when it is sorted.

A particular type of trimmed mean is the *interquartile mean*. 
This is where the bottom and top quartiles are discarded, and the mean is based on the remaining elements.
```{r, echo=T}
iqr_mean <- function(x){
  q1 <- quantile(x, probs = 0.25)
  q3 <- quantile(x, probs = 0.75)
  x[x > q1 & x < q3] %>% 
    mean()
}

iqr_mean(rt_data)
```


An alternative to the trimmed mean, which discards elements, is the *winsorized mean* which replaces values at the extremes of the distributions with values at the thresholds of these extremes.
For example, we could replace elements below the 10th percentile by the value of the 10th percentile, and likewise replace elements above the 90th percentile by the value of the 90th percentile.
The following code implements a winsorized mean function.
```{r, echo=T}
winsorized_mean <- function(x, trim = 0.1){
  low <- quantile(x, probs = trim) # 
  high <- quantile(x, probs = 1 - trim)
  
  x[x < low] <- low
  x[x > high] <- high
  
  mean(x)
}

winsorized_mean(rt_data, trim = 0.1)
```

Another robust measure of central tendency is the *midhinge*, which is the centre point between the first and third quartiles. 
This is equivalent to the arithmetic mean of these two values.
```{r, echo=T}
midhinge <- function(x){
  quantile(x, probs = c(0.25, 0.75)) %>% 
    mean()
}

midhinge(rt_data)
```
The midhinge is a more robust measure of the central tendency than the *midrange*, which is the centre of the full range of the distribution and so equivalent to the the arithmetic mean of its minimum and the maximum values.
```{r, echo=T}
midrange <- function(x){
  range(x) %>% mean() 
}

midrange(rt_data)
```

A related robust measure is the *trimean*, which is defined as the average of the median and the midhinge.
Given that the midhinge is
$$
\textrm{midhinge} = \frac{Q_1 + Q_3}{2},
$$
where $Q_1$ and $Q_3$ are the first and third quartiles, the *trimean* is defined as 
$$
\textrm{trimean} = \frac{Q_2 + \frac{Q_1 + Q_3}{2}}{2} = \frac{Q_1 + 2Q_2 + Q_3}{4}.
$$
As such, we can see the trimean as weighted average of the first, second, and third quartiles.
```{r, echo=T}
trimean <- function(x){
  c(midhinge(x), median(x)) %>% 
    mean()
}

trimean(rt_data)
```

# Measures of dispersion

## Variance and standard deviation

Just as the mean is usually default choice, for good or ill, as a measure of central tendency, the standard measure of the dispersion of distribution is the *variance* or *standard deviation*. 
These two measures should be seen as essentially one measure given that the standard deviation is simply the square root of the variance.

For a continuous random variable $X$, the variance is defined as
$$
\textrm{variance} = \int (x - \bar{x})^2 \Prob{X = x} dx.
$$
While for a discrete random variable $X$, it is defined as 
$$
\textrm{variance} = \sum_{k=1}^K (x_k - \bar{x})^2 \Prob{X = x_k}.
$$
From this, we can see that the variance of a random variable is defined as the expected value of the squared difference of values from the mean. 
In other words, we could state the variance as follows.
$$
\textrm{variance} = \langle d^2 \rangle,\quad \text{where $d = x - \bar{x}$}.
$$
As we saw with the case of the mean, a finite sample of values $x_1, x_2 \ldots x_n$ can be seen as a discrete probability distribution defined at points $x_1, x_2 \ldots x_n$ and each with probability $\frac{1}{n}$.
From this, for a finite sample, the variance is defined as
$$
\textrm{variance} = 
  \sum_{i=1}^n (x_i - \bar{x})^2 \Prob{X = x_i} =
  \sum_{i=1}^n (x_i - \bar{x})^2 \frac{1}{n} =
  \frac{1}{n} \sum_{i=1}^n (x_i - \bar{x})^2.
$$
This is exactly the mean of the squared differences from the mean.
Though R has a built in function for the sample variance, as we will soon see, the sample variance as we have just defined it can be calculated with the following function.
```{r, echo=T}
variance <- function(x){
  mean((x - mean(x))^2)
}
```

One issue with the sample variance as just defined is that it is a *biased* estimator of the variance of the probability distribution of which $x_1, x_2 \ldots x_n$ are assumed to be a sample.
An unbiased estimator of the population's variance is defined as follows.
$$
\textrm{variance} = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2.
$$
This is what is calculated by R's built in variance function `var`. 
Applied to our `rt_data` data, the variance is as follows.
```{r, echo=T}
var(rt_data)
```
The standard deviation is the square root of this number, which is calculated by R's `sd` function, as we see in the following code.
```{r, echo=T}
var(rt_data) %>% sqrt()
sd(rt_data)
```

Given that the variance is exactly, or close to exactly, the average of the squared distances of the values from the mean, to intuitively understand the value of a variance, we must think in terms of squared distances from the mean.
Because the standard deviation is the square root of the variance, its value will be approximately the average distance of all points from the mean. 
For this reason, the standard deviation may be easier to intuitively understand than the variance.
In addition, the standard deviation affords some valuable rules of thumb to understand the distribution of the values of a variable.
For example, for any approximately normally distributed values, approximately 70% of values are within 1 standard deviation from the mean, approximately 95% of values are within 2 standard deviation from the mean, and approximately 99% of values are within 2.5 standard deviations from the mean. 
For other kinds distributions, these rules of thumb may be over or underestimates of the distribution of values. 
However, *Chebyshev's inequality* ensures that no more than $\frac{1}{k^2}$ of the distribution's values will be more than $k$ standard deviations from the mean. 
And so, even in the most extreme cases, 75% of values will always be within 2 standard deviations from the mean, and approximately 90% of cases will be within 3 standard deviations from the mean.

## Trimmed and winsorized estimates of variance and standard deviation

The variance and standard deviation are doubly or triply susceptible to outliers.
They are based on the value of the mean, which is itself prone to the influence outliers.
They are based on squared differences from the mean, which will increase the influence of large differences. 
They are based on a further arithmetic mean operation, which again is unduly influenced by large values. 

Just as with the mean, however, we may trim or winsorize the values before calculating the variance or standard deviation
The following code creates a function that will trim values from the extremes of a sample, as described previously, before applying a descriptive statistic function. 
By default, it uses `mean`, but we can replace this with any function.
```{r, echo=T}
trimmed_descriptive <- function(x, trim = 0.1, descriptive = mean){
  
  n <- length(x)
  lo <- floor(n * trim) + 1
  hi <- n + 1 - lo
  sort(x)[lo:hi] %>% 
    descriptive()
  
}

trimmed_descriptive(rt_data, descriptive = var)
trimmed_descriptive(rt_data, descriptive = sd)
```
We can also define a winsorizing function that can be used with any descriptive statistic.
```{r, echo=T}
winsorized_descriptive <- function(x, trim = 0.1, descriptive = mean){
  low <- quantile(x, probs = trim) # 
  high <- quantile(x, probs = 1 - trim)
  
  x[x < low] <- low
  x[x > high] <- high
  
  descriptive(x)
}

winsorized_descriptive(rt_data, trim = 0.1, var)
winsorized_descriptive(rt_data, trim = 0.1, sd)
```


## Median absolute deviation

A much more robust alternative to the variance and standard deviation is the *median absolute deviation from the median* (\mad).
As the name implies, it is the median of the absolute differences of all values from the median, and so is defined as follows.
$$
\textrm{mad}= \textrm{median}( \vert x_i - m\vert )
$$
We can code this in R as follows.
```{r, echo=T}
sample_mad <- function(x){
  median(abs(x - median(x)))
}

sample_mad(rt_data)
```
The \mad is very straightforward to understand. 
For example, we know immediately that exactly half of all values are less than the value of the \mad from the median.

In the case of the normal distribution, $\textrm{mad} \approx \sigma/1.48$, where $\sigma$ is the distribution's standard deviation.
Given this, the \mad is often scaled by approximately $1.48$ so as to act as a robust estimator of the standard deviation. 
In R, the function the built-in command, part of the `stats` package, for calculating the \mad is `mad` and this is by default calculated as follows.
$$
\textrm{mad}= 1.4826 \times \textrm{median}( \vert x_i - m\vert )
$$
We can verify this in the following code.
```{r, echo=T}
1.4826 * sample_mad(rt_data)
mad(rt_data)
```
Note how much lower the \mad is compared to the sample standard deviation, which is $\approx `r round(sd(rt_data),2)`$

## Range estimates of dispersion

By far the simplest measure of the dispersion of a set of values is the *range*, which the difference between the maximum and minimum values. 
In R, the command `range` returns the minimum and the maximum values per se, rather than their difference, but we can calculate the range easily as in the following example.
```{r, echo=T}
max(rt_data) - min(rt_data)
```
Although the range is informative, it is also obviously extremely prone to undue influence of outliers given that is defined in terms of the extremities of the set of values.
More robust alternative estimates of the dispersion are based on quantiles. 
The following functions returns the difference between some specified upper and lower quantile value.
```{r, echo=T}
quantile_range <- function(x, lower, upper){
  quantile(x, probs = c(lower, upper)) %>%
    unname() %>% 
    diff()
}
```
Given that the minimum and maximum values of a set of numbers are the \nth{0} and the \nth{100} percentiles, respectively, this function can be used to return the standard defined range, as in the following example.
```{r, echo=T}
quantile_range(rt_data, lower = 0.0, upper = 1.0)
```
This is obviously the 100% inner range of the values.
The 90% inner range of `rt_data` is as follows.
```{r, echo=T}
quantile_range(rt_data, lower = 0.05, upper = 0.95)
```


The range from the \nth{10} to the \nth{90} percentile, which gives the 80% inner range, is as follows.
```{r, echo=T}
quantile_range(rt_data, lower = 0.1, upper = 0.9)
```
This range is known as the *interdecile range*.
On the other hand, the range from the \nth{25} to the \nth{75} percentile, which gives the 50% inner range, is as follows.
```{r, echo=T}
quantile_range(rt_data, lower = 0.25, upper = 0.75)
```
This range is known as the *interquartile range* (\iqr), which can also be calculated using the built-in `IQR` command in R.
```{r, echo=T}
IQR(rt_data)
```
Just as with \mad, in normal distributions, there is a constant relationship between the \iqr and the standard deviation.
Specifically, $\textrm{iqr} \approx 1.349 \times \sigma$, and so $\textrm{iqr} / 1.349$ is a robust estimator of the standard deviation.
```{r, echo = T}
IQR(rt_data) / 1.349
```

# Measure of skewness

Skewness is a measure of the asymmetry of a distribution of numbers.
For a random variable $X$, the skewness is the *third standardized moment* defined as follows.
$$
\textrm{skew} = \frac{\left\langle (X - \mu)^3 \right\rangle}{
                      \left\langle (X - \mu)^2 \right\rangle^{3/2}} =
\frac{\left\langle (X - \mu)^3 \right\rangle}{\sigma^{3}}
$$
The numerator is the expected value of the *third central moment*, which is the expected value of the cube of the difference of the variable from the mean, i.e.,
$$
\left\langle (X - \mu)^3 \right\rangle = \int (x - \mu)^3\ \Prob{X = x} dx.
$$
The denominator is the cube of the standard deviation given that the standard deviation $\sigma$ is the square root of second central moment $\left\langle (X - \mu)^2 \right\rangle$.

Whenever there is a longer tail to the right of the distribution, the skewness takes a positive value. 
Thus, *right skewed* or *right tailed* distributions have *positive skew*.
By contrast, whenever there is a longer tail to the left of the distribution, the skewness takes a positive value, and so *left skewed* or *left tailed* distributions have *negative skew*.

In a finite sample of $n$ values, the skewness is calculated as follows:
$$
\textrm{skew} = \frac{\frac{1}{n}\sum_{i=1}^n (x_i - \bar{x})^3}{s^3},
$$
where $\bar{x}$ and $s$ are the sample mean and sample standard deviation, respectively.
In R, we can implement this function as follows.
```{r, echo=T}
skewness <- function(x, dof=1){
  xbar <- mean(x)
  s <- sd(x)
  mean((x - xbar)^3)/s^3
}

skewness(rt_data)
```
There is no built-in function for skewness in R.
However, the function just defined is also available as `skew` in the `psych` package.
```{r, echo=T}
psych::skew(rt_data)
```
A slight variant is also available as `skewness` in the package `moments`.
```{r, echo=T}
moments::skewness(rt_data)
```
In this version, the standard deviation is calculated based on a denominator of $n$ rather than $n-1$.


### Trimmed and winsorized skewness

The measure of sample skewness just given is highly sensitive to outliers. 
This is so because it is based on sample means and standard deviations, and also because it involves cubic functions. 
We can, however, apply this function to trimmed or winsorized samples.
We may use the above defined `trimmed_descriptive` and `winsorized_descriptive` with the `skewness` function as in the following example.
```{r, echo=T}
trimmed_descriptive(rt_data, descriptive = skewness)
winsorized_descriptive(rt_data, descriptive = skewness)
```
Notice how these measures are now both negative and with much lower absolute values.


### Quantile skewness

In a symmetric distribution, the median would be in the exact centre of any quantile range.
For example, in a symmetric distribution, the median would lie in the centre of the interval between the lower and upper quartiles, and likewise in the centre of the interdecile range and so on.
We can use this fact to get a measure of asymmetry. 
If the median is closer to the lower quantile than the corresponding upper quantile, the distribution is right tailed and so there is a positive skew. 
Conversely, if it closer to the upper quantile than the corresponding lower one, the distribution is left tailed and there is a negative skew.
This leads to the following definition of *quantile skewness*.
$$
\textrm{skew}_{Q} = \frac{(Q_u - m) - (m - Q_l)}{Q_u - Q_l},
$$
where $Q_u$ and $Q_l$ are the upper and lower quantiles, respectively, and $m$ is the median.
We can implement this as follows.
```{r, echo=T}
qskewness <- function(x, p=0.25){
  Q <- quantile(x, probs = c(p, 0.5, 1 - p)) %>% 
    unname()
  Q_l <- Q[1]; m <- Q[2]; Q_u <- Q[3]
  ((Q_u - m) - (m - Q_l)) / (Q_u - Q_l)
}

qskewness(rt_data) # quart_dataile skew 
qskewness(rt_data, p = 1/8) # octile skew
qskewness(rt_data, p = 1/10) # decile skew
```
Notice how these values are in line with those the trimmed and winsorized skewness measures.

### Nonparametric skewness

The following function is known as the *nonparametric skewness* measure.
$$
\textrm{skew} = \frac{\bar{x} - m}{s},
$$ 
where $\bar{x}$ and $s$ are the sample mean and sample standard deviation as before, and $m$ is the same median.
It will always be positive if the mean to the right of the median, and negative if the mean is to the left of the median, and zero if the mean and median are identical.
It is also bounded between $-1$ and $1$, given that the median is always less than 1 standard deviation from the mean.
It should be noted, however, that its values do not correspond to those of skewness functions defined above.
It is easily implemented as follows.
```{r, echo=T}
npskew <- function(x){
  (mean(x) - median(x))/sd(x)
}

npskew(rt_data)
```

*Pearson's second skewness coefficient* is defined as 3 times the nonparametric skewness measure. 
$$
\textrm{skew} = 3 \left(\frac{\bar{x} - m}{s}\right),
$$
This can be easily implemented as follows.
```{r, echo=T}
pearson_skew2 <- function(x){
  3 * npskew(x)
}

pearson_skew2(rt_data)
```
Like the nonparametric skewness measure, it will be positive if the mean is greater than the median, negative if the mean is less than the median, zero if they are identical. 
It will also be bounded between $-3$ and $3$.

# Measures of kurtosis

Kurtosis is often described as measuring how *peaky* a distribution is. 
However, this is a misconception and kurtosis is better understood as relating to the heaviness of a distribution's tails.
@westfall2014kurtosis argues that this misconception of kurtosis as pertaining to nature of a distribution's peak is due to Karl Pearson's suggestions that estimates of kurtosis could be based on whether a distribution is more, or less, or equally *flat-topped* compared to a normal distribution. 
By contrast, @westfall2014kurtosis shows that it is the mass in the tails of the distribution that primarily defines the kurtosis of a distribution.

In a random variable $X$, kurtosis is defined as the *fourth standardized moment*:
$$
\textrm{kurtosis} = \frac{\left\langle (X - \mu)^4 \right\rangle}{
                      \left\langle (X - \mu)^2 \right\rangle^{2}} =
\frac{\left\langle (X - \mu)^4 \right\rangle}{\sigma^{4}}.
$$
The sample kurtosis is defined analogously as follows.
$$
\textrm{kurtosis} = 
\frac{\frac{1}{n} \sum_{i=1}^4 (x_i - \bar{x})^4}{s^{4}},
$$
where $\bar{x}$ and $s$ are the mean and standard deviation.
This simplifies to the following.
$$
\textrm{kurtosis} = \frac{1}{n} \sum_{i=1}^n z^4,
$$
where $z_i = (x_i - \bar{x})/s$.
This function can be implemented easily in R as follows.
```{r, echo=T}
kurtosis <- function(x){
  z <- (x - mean(x))/sd(x)
  mean(z^4)
}

kurtosis(rt_data)
```
This function is also available from `moments::kurtosis`, though in that case, the standard deviation is calculated with a denominator of $n$ rather than $n-1$.

In a normal distribution, the kurtosis, as defined above has a value of 3.0.
For this reason, it is conventional to subtract 3.0 from the kurtosis function, both the population and sample kurtosis functions.
This is properly known as *excess kurtosis*, but in some implementations, it is not always clearly stated that the excess kurtosis rather than kurtosis per se is being calculated.
Here, we will explicitly distinguish between these two functions, and so the sample excess kurtosis is defined as follows.
```{r, echo=T}
excess_kurtosis <- function(x){
  kurtosis(x) - 3
}

excess_kurtosis(rt_data)
```

Let us look at the excess kurtosis of a number of familiar probability distributions.
```{r, echo=T}
N <- 1e4
distributions <- list(normal = rnorm(N),
                      t_10 = rt(N, df = 10),
                      t_7 = rt(N, df = 7),
                      t_5 = rt(N, df = 5),
                      uniform = runif(N)
)
map_dbl(distributions, excess_kurtosis) %>% 
  round(digits = 2)
```
The function `rnorm` samples from a normal distribution. 
As we can see, the sample excess kurtosis of the normal samples are close to zero as expected.
The function `rt` samples from a Student t-distribution.
The lower the degrees of freedom, the heavier the tails. 
Heavy tailed distribution have additional mass in their tails and less in their centres compared to a normal distribution.
As we can see, as the tails get heavier, the corresponding sample kurtosis increases.
The `runif` function samples from a uniform distribution.
Uniform distributions have essentially no tails; all the mass in the centre.
As we can see, the excess kurtosis of this distribution is negative.
Distributions with zero or close to zero excess kurtosis are known as *mesokurtic*.
Distributions with positive excess kurtosis are known as *leptokurtic*.
Distribution with negative excess kurtosis are known as *platykurtic*.

Given that kurtosis is primarily a measure of the amount of mass in the tails of the distribution relative to the mass in its centre, it is highly sensitive to the values in the extreme of a distribution.
Were we to trim or winsorize the tails, as we have done above for the calculation of other descriptive statistics, we would necessarily remove values from the extremes of the sample.
This could drastically distort the estimate of the kurtosis. 
To see this, consider the values of excess kurtosis for the `r length(distributions)` samples in the `distributions` data set above after we have trimmed and winsorized 10% of values from both extremes of each sample. 
Note that here we are using the `map_dbl` function from `purrr`, which is part of the `tidyverse`.
This will apply the trimmed or winsorized functions to each element of `distributions`.
We will explore `map_dbl` in Chapter 6.
```{r, echo=T}
map_dbl(distributions, 
        ~trimmed_descriptive(., descriptive = excess_kurtosis)) %>% 
  round(digits = 2)

map_dbl(distributions, 
        ~winsorized_descriptive(., descriptive = excess_kurtosis)) %>% 
  round(digits = 2)
```
With the exception of the platykurtic `uniform` distribution, the estimates of the excess kurtosis of the other distributions have been distorted to such an extent that the mesokurtic and leptokurtic distributions now appear to platykurtic.

This raises the question of what exactly is an outlier. 
Rules of thumb such as that outliers are any values beyond 1.5 times the interquartile range above the third quartile or below the first quartile will necessarily distort leptokurtic distributions if the values meeting this definition were removed.
The presence of values beyond these limits may not indicate anomalous results per se but simply typical characteristic values of heavy tailed distributions.
What defines an outlier then can only be determined after some assumptions about the true underlying distribution have been made. 
For example, if we assume the distribution is mesokurtic and roughly symmetrical, then values beyond the interquartile defined limits just mentioned can be classified as anomalous.
In general then, classifying values as outliers is a challenging problem involving assumptions about the distributions then the probabilistic modelling of the data based on this assumptions. 
Short of taking these steps, we must be very cautious in how we define classifiers, especially when calculating quantities such as kurtosis. 

In the following code, we create a functions to remove and winsorize outliers.
The outliers are defined as any values beyond `k`, defaulting to `k = 1.5`, times the interquartile range above or below, respectively, the \nth{3} and \nth{1} quartiles.
```{r, echo=T}
trim_outliers <- function(x, k = 1.5){
  iqr <- IQR(x)
  limits <- quantile(x, probs = c(1, 3)/4) + c(-1, 1) * k * iqr
  x[(x > limits[1]) & (x < limits[2])]
}

winsorize_outliers <- function(x, k = 1.5){
  iqr <- IQR(x)
  limits <- quantile(x, probs = c(1, 3)/4) + c(-1, 1) * k * iqr
  outlier_index <- (x < limits[1]) | (x > limits[2])
  x[outlier_index] <- median(x)
  x
}
```
If we set `k` to a value greater than 1.5, we will be more stringent in our definition of outliers, and so there is considerably less distortion to the estimates of excess kurtosis.
```{r, echo=T}
map_dbl(distributions, 
        ~excess_kurtosis(trim_outliers(., k = 3.0))
)
map_dbl(distributions, 
        ~excess_kurtosis(winsorize_outliers(., k = 3.0))
)
```

```{r}
anomaly <- rt_data[2]
iqr <- IQR(rt_data)
anomaly_dist <- (anomaly - quantile(rt_data, probs = 0.75))/iqr
```


Our `rt_data` has an extraordinarily high excess kurtosis value of `r round(excess_kurtosis(rt_data), digits = 3)`.
However, the value of `r rt_data[2]` is `r round(anomaly_dist, 2)` times the interquartile range above the \nth{3} quartile. 
When we remove this value, or replace it with the median value, the kurtosis changes dramatically. 
```{r, echo=T}
excess_kurtosis(rt_data[-2])

rt_replace <- rt_data
rt_replace[2] <- median(rt_data)
excess_kurtosis(rt_replace)
```

### Quantile based measures of kurtosis

As with other descriptive statistics, we can use quantiles to allow us to calculate robust estimates of kurtosis.
One simple procedure us to calculate the ratio of the 95%, or 99% etc, inner quantile interval to the inter quartile range. 
We can denote this as follows.
$$
\textrm{qkurtosis}_{p} = \frac{Q_{1-p} - Q_{p}}{Q_3 - Q_1},
$$
where $Q_1$ and $Q_3$ are the first and third quartiles, respectively, and $Q_{1-p}$ and $Q_p$ are the quantiles at $1-p$ and $p$, respectively, where $0 \leq p \leq 1$.
Clearly, when $p=0.025$, $\textrm{qkurtosis}_{p}$ is the ratio of the 95% inner quantile interval to the interquartile range. 

In a normal distribution, for example, that value of $\textrm{qkurtosis}_{p=0.025}$ is approximately `r round(abs(qnorm(0.025))/abs(qnorm(0.25)), 2)`, and the value of $\textrm{qkurtosis}_{p=0.005}$ is approximately `r round(abs(qnorm(0.005))/abs(qnorm(0.25)), 2)`.
In leptokurtic distributions, these values are higher.
For example, in the t-distribution with $\nu = 5$ degrees of freedom, the values of
 $\textrm{qkurtosis}_{p=0.025}$ and $\textrm{qkurtosis}_{p=0.005}$ are 
  `r round(abs(qt(0.025, df=5))/abs(qt(0.25, df=5)), 2)` and  `r round(abs(qt(0.005, df=5))/abs(qt(0.25, df=5)), 2)`, respectively.
On the other hand, in platykurtic distributions, these values are lower. 
For example, in a uniform distribution they are `r round(0.95/0.5, 2)` and `r round(0.99/0.5, 2)`, respectively.

It should be noted, and we see below, that the values of $\textrm{qkurtosis}_{p}$ and of the standard $\textrm{kurtosis}$ function will not be identical or even linearly related, but there will be a clear monotonic correspondence between them in most cases. 
Also, if we subtract 3 from the value of $\textrm{qkurtosis}_{p}$, we will get a quantile based counterpart of excess kurtosis.

We can also extend this quantile estimate of kurtosis as follows.
$$
\textrm{qkurtosis}_{u,p} = \frac{Q_{1-p} - Q_{p}}{Q_{1-u} - Q_{u}},
$$
where $0 \leq p \leq u \leq 1$.
If $u = 0.25$, then $\textrm{qkurtosis}_{u,p}$ is identical to $\textrm{qkurtosis}_{p}$ defined above.

We can implement these functions in R as follows.
```{r, echo=T}
qkurtosis <- function(x, u=0.25, p=0.025){
  Q <- quantile(x, probs = c(p, u, 1-u, 1-p))
  diff(Q[c(1, 4)])/diff(Q[c(2, 3)])
}
```

Applying this function to the `distributions` data sets, we can investigate the effect of using different values of $u$ and $p$.
```{r, echo=T}
map_dbl(distributions, qkurtosis) - 3 

map_dbl(distributions, 
        ~qkurtosis(., p = 0.01/2)
) - 3 

map_dbl(distributions, 
        ~qkurtosis(., u=0.2, p = 0.01/2)
) - 3 
```
Clearly, none of these exactly estimates match the kurtosis estimates using the standard kurtosis function. 
However, in all cases, there is clear correspondence between the two.

The $\textrm{qkurtosis}_{u,p}$ is very straightforward to explain or understand. 
It gives the ratio of the width of portion of the distribution containing most of the probability mass to the width of its centre.
If most of the mass is near the centre, as is the case with platykurtic distributions, this ratio will be relatively low. 
If the tails contain a lot of mass, as is the case with leptokurtic distributions, this ratio will be relatively high.

# Graphical exploration of univariate distributions

Thus far, we have used different methods for measuring characteristics of samples from univariate distributions such as their location, scale, and shape.
Each of these methods provides us with valuable perspectives on the sample and consequently also provide insights into the nature of the distribution from which we assume this sample has been drawn.
These are essential preliminary steps before we embark on more formal modelling of these distributions. 
In addition to these methods, there are valuable graphical methods for exploring univariate distributions. 
These ought to be seen as complementary to the quantitative methods described above.
We have already seen a number of relevant methods, such as histograms and boxplots, in Chapter 4.
Here, we will primarily focus on methods not covered in Chapter 4.

## Stem and leaf plots

One of the simplest possible graphical methods for exploring a sample of univariate values is the *stem and leaf* plot, first proposed by @tukey1977exploratory.
In the built-in `graphics` package in R, the function `stem` produces stem and leaf plots.
In the following code, we apply `stem` to a set of 6 example values.
```{r, echo=T, eval=F}
c(12, 13, 13, 16, 21, 57) %>% 
  stem()
```
```{r, comment=''}
c(12, 13, 13, 16, 21, 57) %>% 
  stem()
```
To the left of the `|` are the *stems*. 
These are the leading digits of the set of values.
To the right of `|` are the *leaves*. 
These are the remaining digits of the values. 
For example, `1 | 2336` indicates that we have 4 values whose first digit is $1$ and whose remaining digits are $2$, $3$, $3$, and $6$.
In a stem and leaf plot, we can often display all the values in the sample, and thus there is not data loss or obscuring of the data, and yet we can still appreciate major features of the distribution such as the location, scale, and shape. 

The command `stem.leaf` from the `aplpack` package provides a classic Tukey style version of the stem and leaf plot.
In the following code, we apply the `stem.leaf` command to the `rt_data`.
```{r, echo=T, eval=F}
library(aplpack)
stem.leaf(rt_data, trim.outliers = FALSE, unit=10)
```
```{r, comment=''}
aplpack::stem.leaf(rt_data, trim.outliers = FALSE, unit=10)
```
Here, we have specified that outliers should not be trimmed and also that the order of magnitude of the leaves should be $10$.
As the legend indicates, rows such as `2 | 5` indicate a value on the order of $250$, while `3 | 1156` indicates two values on the order of $310$ (in this case, these are both the $317$), one value on the order of $350$ (in this case, this is $350$), and one on the order of $360$ (this is $367$).
The first column of numbers indicate the number of values at or more extreme than the values represented by the corresponding stem and leaf. 
For example, the `5` on the second row indicates that there are 5 values that as or more extreme, relative to the median value, than the values represented by `3 | 1156`.
In this column, the number in round brackets, i.e., `(8)` indicates that the median value is in the set of values represented by `5 | 01166688`.

## Histograms

Stem and leaf plots are only useful for relatively small data sets.
When the number of values becomes relatively large, the number of digits in the leaves and hence their lengths can become excessive. 
If we decrease the units of the plot, and thus decrease its granularity, we may end up excessive numbers of rows; too many to display even on a single page.
In this situation, it is preferable to use a histograms.
We have already described how to make histograms using `ggplot2` in Chapter 4.
Here, we provide just a brief recap example.

For this, we will use the example of the per capita gross domestic product (GDP) in a set of different countries.
```{r, echo=T}
gdp_df <- read_csv('data/nominal_gdp_per_capita.csv')
```
As a first step, we will visualize this data with a histogram using the following code, which is shown in Figure \ref{fig:gdp_histogram}.
```{r, gdp_histogram, fig.align='center', fig.cap='The distribution of the per capita gross domestic product in a set of different countries.', echo=T, out.width='0.667\\textwidth'}
gdp_df %>%
  ggplot(aes(x = gdp)) +
  geom_histogram(col='white', binwidth = 2500) +
  xlab('Per capita GDP in USD equivalents')
```

As with any histogram, we must choose a bin width or the number of bins into which we will divide all the values. 
Although `ggplot` will default to usually 30 bins in its histograms, it will raise a warning to explicitly pick a better number value. 
In this case, we will choose a `binwidth` of 2500.

Even with this simple visualization, we have learned a lot.
We see that most countries have per capita GDP lower than \$30,000, and in fact most of these countries have incomes lower than around \$15,000.
The distribution is spread out from close to zero to close to \$100,000, and it is is highly asymmetric, with a long tail to the right.


## Boxplots

Another useful and complementary visualization method is the Tukey boxplot.
Again, we have described how to make boxplots with `ggplot2` in detail in Chapter 4, and so only provide a recap here.

A horizontal Tukey boxplot with all points shown as *jittered* points, and with a *rug plot* along the horizontal axis, can be created with the following code, which is shown in Figure \ref{fig:gdp_box}.
```{r gdp_box, fig.align='center', fig.cap='A Tukey boxplot of the distribution of per capita GDP across a set of countries.', out.width='\\textwidth', echo=T}
ggplot(gdp_df,
       aes(x = '', y = gdp)
) + geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.25, size = 0.5, alpha = 0.5) + 
  geom_rug(sides = 'l') +
  coord_flip() +
  theme(aspect.ratio = 1/5) +
  xlab('')
```
As we have seen in Chapter 4, the box itself in the boxplot gives us the 25th, 50th, and 75th percentiles of the data.
From this, we see more clearly that 50% of the countries have per capita incomes lower than around \$6000 and 75% of countries have incomes lower than around \$17,000. 
The top 25% of the countries are spread from around \$17,000 to above \$100,000.


## Empirical cumulative distribution functions

The *cumulative distribution function*, or just simply the *distribution function*, of a random variable $X$ is defined as follows.
$$
F(t) = \Prob{X < t} = \int_{\infty}^t \Prob{X = x} dx.
$$
In other words, $F(t)$ is the probability that the random variable takes on a value of less than $t$.
For a sample of values $x_1, x_2 \ldots x_n$, we define the *empirical cumulative distribution function* (\ecdf).
$$
F(t) = \frac{1}{n}\sum_{i=1}^n I_{x_i \leq t},
$$
where $I_{A}$ takes the value of 1 if its argument $A$ is true and takes the value of zero otherwise.
We can plot the \ecdf of a sample with `ggplot` as follows.
```{r, gdp_ecdf, fig.align='center', fig.cap='The \\ecdf of the per capita GDP across a set of countries.', out.width='0.75\\textwidth', echo=T}
gdp_df %>% 
  ggplot(aes(x = gdp)) + 
  stat_ecdf()
```
From the \ecdf, by using the percentiles on the y-axis and find the values of `gdp` on the x-axis that they correspond to, we can see that most of the probability mass is concentrated as low values of `gdp`, and that there is a long tail to the right. 

## Q-Q and P-P plots

Distribution functions, both theoretical and empirical, are used to make so-called Q-Q and P-P plots.
Both Q-Q and P-P plots are graphical techniques that are used to essentially compare distribution functions. 
While both of these techniques are widely used, especially to compare a sample to a theoretical distribution, the meaning of neither technique is particularly self-evident to the untrained eye.
It is necessary, therefore, to understand the technicalities of both techniques before they can used in practice.

To understand both Q-Q and P-P plots, let us start with two random variables $X$ and $Y$ that have density functions $f(x)$ and $g(y)$, respectively, and cumulative distribution functions $F(x)$ and $G(y)$, respectively.
In principle, for any given quantile value $p \in (0, 1)$, we can calculate $x = F^{-1}(p)$ and $y = G^{-1}(p)$, where $F^{-1}$ and $G^{-1}$ are the inverses of the cumulative distribution functions, assuming that $F^{-1}$ and $G^{-1}$ exist.
Therefore, the pair $(x, y)$ are values of the random variables $X$ and $Y$, respectively, that correspond to the same quantile value. 
If, for a range of quantile values $0 < p_1 < p_2 \ldots < p_i < \ldots p_n$, we calculate $x_i = F^{-1}(p_i)$ and $y_i = G^{-1}(p_i)$ and then plot each pair $(x_i, y_i)$, we produce a Q-Q plot. 
If the density functions $f(x)$ and $g(y)$ are identical, the the points in the Q-Q plot will fall on the identity line, i.e. the line through the origin with a slope of $1$.
If $f(x)$ and $g(x)$ differ in their location only, i.e. one is a shifted version of the other, then the $(x_i, y_i)$ will still fall on a straight line with slope equal to $1$ whose intercept is no longer zero and represent how much the mean of $Y$ is offset relative to that of $X$.
If $f(x)$ and $g(x)$ have the same location but differ in their scale, then the $(x_i, y_i)$ will again fall on a straight line, but the slope will not equal $1$ but the intercept will be zero.  
The slope represents the ratio of standard deviation of $Y$ to that of $X$.
If $f(x)$ and $g(x)$ differ in both location and their scale, then the $(x_i, y_i)$ will yet again fall on a straight line, but the slope will not equal $1$ and the intercept will not be zero.  
These four scenarios are illustrated with normal distributions in Figure \ref{fig:qq_examples}.

```{r, qq_examples, fig.align='center', fig.cap='Q-Q plots corresponding to (a) two identical normal distributions, (b) two normal distributions that differ in their mean only, (c) two normal distributions that differ in the standard deviation only, and (d) two normal distributions that differ in the means and their standard deviations.', out.width='0.75\\textwidth'}
qq_norm_plot <- function(mu=c(0, 0), sigma=c(1, 1), N=25, epsilon=1e-2){
  epsilon <- 0.01
  p <- seq(epsilon, 1-epsilon, length.out = N)
  tibble(x = qnorm(p, mean = mu[1], sd = sigma[1]),
         y = qnorm(p, mean = mu[2], sd = sigma[2])
  ) %>% ggplot() + 
    geom_abline(intercept = 0, slope = 1, colour = 'red') +
    geom_point(aes(x = x, y = y))
}

P_1 <- qq_norm_plot()
P_2 <- qq_norm_plot(mu = c(0, 1))
P_3 <- qq_norm_plot(sigma = c(1, 2))
P_4 <- qq_norm_plot(mu = c(0, 1), sigma = c(1, 2))

plot_grid(P_1, P_2, P_3, P_4, 
          labels = c('a', 'b', 'c', 'd'),
          nrow = 2)
```

```{r, qq_examples_1, fig.align='center', fig.cap='Q-Q plots comparing a normal distribution represented by the x-axis to four alternative distributions. In a) the normal distribution is compared to a t-distribution with $\nu = 5$ degrees of freedom. In (b), we compare the normal distribution to a t-distribution with $\nu = 3$. In (c), the normal distribution is compared to a $\\chi^2$ distribution with $\nu = 1$. In (d), the normal distribution is compared to an exponential distribution with rate parameter $\\lambda = 1$. In all cases, the red line depicts the identity line.', out.width='0.75\\textwidth'}
qq_nonnorm_plot <- function(F_inv = qnorm, 
                            G_inv = function(p) qt(p, df = 3), 
                            N=25, 
                            epsilon=1e-2){
  epsilon <- 0.01
  p <- seq(epsilon, 1-epsilon, length.out = N)
  tibble(x = F_inv(p),
         y = G_inv(p)
  ) %>% ggplot() + 
    geom_abline(intercept = 0, slope = 1, colour = 'red') +
    geom_point(aes(x = x, y = y)) + geom_line(aes(x = x, y = y), size = 0.5, alpha = 0.5)
}

P_1 <- qq_nonnorm_plot(F_inv = qnorm, 
                       G_inv = function(p) qt(p, df = 5))
P_2 <- qq_nonnorm_plot(F_inv = qnorm, 
                       G_inv = function(p) qt(p, df = 3))
P_3 <- qq_nonnorm_plot(F_inv = qnorm, 
                       G_inv = function(p) qchisq(p, df = 1))
P_4 <- qq_nonnorm_plot(F_inv = qnorm, 
                       G_inv = function(p) qexp(p, rate = 1))

plot_grid(P_1, P_2, P_3, P_4, 
          labels = c('a', 'b', 'c', 'd'),
          nrow = 2)
```



Whenever distributions differ only in terms of their location and scale, then their Q-Q plots will fall on a straight line and it will be possible to see how their locations and scales differ from one another. 
Two distributions that differ in their shape will correspond to Q-Q plots where the points do not fall on straight lines. 
To illustrate this, in Figure \ref{fig:qq_examples_1}, we compare standard normal distributions to t-distributions (Figure \ref{fig:qq_examples_1}a-b), a $\chi^2$ distribution (Figure \ref{fig:qq_examples_1}c), and an exponential distribution (Figure \ref{fig:qq_examples_1}d).
In the case of the two t-distributions, we can see that both the left and right tail are more spread out than the corresponding tails in the normal distribution, while the centres are more similar.
In the case of the $\chi^2$ and exponential distributions, we see the left tails are far more concentrated than the left tail of normal distribution, while the right tails are for more spread out.

We may now use Q-Q plots to compare the \ecdf of a sample against the distribution function of a known probability distribution. 
As an example, let us again consider the `rt_data` data.
The quantiles that these values correspond to can be found using R's `ecdf` function.
This function returns a function that can be used to calculate the \ecdf's value for any given value.
```{r, echo=T}
F <- ecdf(rt_data)
(p <- F(rt_data))
```
Before, we proceed, we see here that we have a quantile value of $`r p[2]`$ in the `p` vector.
This will correspond to $\infty$ in any normal distribution.
We can avoid this by first clipping the vector of quantiles so that its minimum is $\epsilon > 0$ and its maximum is $1-\epsilon$.
```{r, echo=T}
clip_vector <- function(p, epsilon=1e-3){
  pmax(pmin(p, 1-epsilon), epsilon)
}

p <- clip_vector(p)
```

We now find the values of a theoretical probability distribution that correspond to these quantiles.
In this case, rather than comparing to a standard normal, we will use a normal distribution with a mean and standard deviation equal to a robust measure of the location and scale of `rt_data`.
For this, we will use its median and \mad, respectively.
We may find these by using the inverse of the cumulative distribution function of that probability, assuming it exists.
For example, the values of standard normal distribution that correspond to these quantiles can be calculated using the `qnorm` function.
```{r, echo=T}
qnorm(p, mean = median(rt_data), sd = mad(rt_data))
```


Now we can plot `rt_data` against `qnorm(p)` to produce our Q-Q plot.
The code for this is below and the plot is shown in Figure \ref{fig:rt_qq}.
```{r rt_qq, fig.align='center', fig.cap='A Q-Q plot of \\texttt{rt\\_data} against a normal distribution whose mean and standard deviation match its location and scale. The red line is the identity line.', out.width='0.67\\textwidth', echo=T}
tibble(y = rt_data,
              x = qnorm(p, mean = median(rt_data), sd = mad(rt_data))
) %>% ggplot(aes(x = x, y = y)) + geom_point() +
  geom_abline(intercept = 0, slope = 1, col='red')
```
We can see here that points in the middle fall roughly on the identity line, but there is more somewhat more spread on the left tail, and an extreme outlier on the right tail.

In Figure \ref{fig:gdp_qq}, we provide the Q-Q plot of the `gdp` values from `gdp_df` against a normal distribution whose mean and standard deviation roughly match the location and scale of `gdp`.
From this, we can see that the values on the left tail are extremely concentrated and the values on the right tail are extremely spread out.
```{r gdp_qq, fig.align='center', fig.cap='A Q-Q plot of the \\texttt{gdp} data against a normal distribution whose mean and standard deviation match its location and scale. The red line is the identity line.', out.width='0.67\\textwidth'}
gdp_df %>% 
  mutate(p = ecdf(gdp)(gdp),
         x = qnorm(p, mean = median(gdp), sd = mad(gdp))
  ) %>% ggplot(aes(x = x, y = gdp)) + 
        geom_point() + 
        geom_abline(intercept = 0, slope = 1, col='red')
```


P-P plots are related to Q-Q plots, but are less well known or widely used.
To understand them, again let us return to the case of the random variables $X$ and $Y$ that have density functions $f(x)$ and $g(y)$, respectively, and cumulative distribution functions $F(x)$ and $G(y)$, respectively.
For any value in $z$, we may calculate $F(z)$ and $G(z)$.
If, for each value in the sequence $z_1 < z_2 \ldots < z_i < \ldots z_n$, we calculate $F(z_i)$ and $G(z_i)$ and then plot each pair $(F(z_i), G(z_i))$, we produce a P-P plot.

In Figure \ref{fig:pp_examples_1}, we provides four P-P plots each comparing different distributions to a normal distribution. 
In Figure \ref{fig:pp_examples_1}a, we compare the t-distribution with $\nu = 1$ degrees of freedom to a standard normal distribution. 
Here, we see that the percentiles in the t-distribution at or below the \nth{10} percentile or at or above the \nth{90} percentile correspond to close to the \nth{0} and \nth{100} percentiles, respectively, in a normal distribution.
This shows that the tails of the t-distribution are more spread out than that of the normal distribution.
By contrast, in Figure \ref{fig:pp_examples_1}d, for example, we compare a uniform from $-1$ to $1$ to a standard normal distribution. 
In this case, we see that the percentiles in the normal distribution at or below the \nth{10} percentile or at or above the \nth{90} percentile correspond to close to the \nth{0} and \nth{100} percentiles, respectively, in the uniform distribution.
This is clearly because the uniform distribution from $-1$ to $1$ is more compact than the normal distribution.

```{r, pp_examples_1, fig.align='center', fig.cap='P-P plots comparing a standard normal distribution represented by the x-axis to four alternative distributions. In a) the normal distribution is compared to a t-distribution with $\nu = 1$ degrees of freedom. In (b), we compare the normal distribution to $\\chi^2$ distribution with $\nu = 3$ degrees of freedom. In (c), the normal distribution is compared to a uniform distribution from $-5$ to $5$. In (d), the normal distribution is compared to a unifrom distribution from $-1$ to $1$. In all cases, the red line depicts the identity line.', out.width='0.75\\textwidth'}
pp_nonnorm_plot <- function(Fd = pnorm, 
                            Gd = function(z) pt(z, df = 3), 
                            N=25, 
                            zlim = c(-5, 5)){
  epsilon <- 0.01
  z <- seq(zlim[1], zlim[2], length.out = N)
  tibble(x = Fd(z),
         y = Gd(z)
  ) %>% ggplot() + 
    geom_abline(intercept = 0, slope = 1, colour = 'red') +
    geom_point(aes(x = x, y = y)) + geom_line(aes(x = x, y = y), size = 0.5, alpha = 0.5)
}
zlim = c(-5, 5)
P_1 <- pp_nonnorm_plot(Fd = pnorm, 
                       Gd = function(z) pt(z, df = 1))
P_2 <- pp_nonnorm_plot(Fd = pnorm, 
                       Gd = function(z) pchisq(z, df = 3))
P_3 <- pp_nonnorm_plot(Fd = pnorm, 
                       Gd = function(z) punif(z, min=zlim[1], max=zlim[2]))
P_4 <- pp_nonnorm_plot(Fd = pnorm, 
                       Gd = function(z) punif(z, min=-1, max=1))

plot_grid(P_1, P_2, P_3, P_4, 
          labels = c('a', 'b', 'c', 'd'))
```


# References
