---
title: "Chapter 2: Introduction to R"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
bibliography:
  - "`r system('kpsewhich mjandrews.bib', intern=TRUE)`"
biblio-style: apalike   
header-includes:
  - \input{header.tex}
---




```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(scipen=9)
set.seed(42)

library(dplyr)
library(stringr)

available_packages <- available.packages(repos = "http://cran.us.r-project.org") %>% as_tibble()

open_licences <- str_detect(available_packages$License, 
                            regex('gpl|bsd|mit|apache|cc|GNU|artistic', ignore_case = T)) %>% 
  mean()
```

# What is R and why should you care

* R is a program for doing statistics and data analysis.
* R's advantages or selling points relative to other programs (e.g, SPSS, SAS, Stata, Minitab,
Python, Matlab, Maple, Mathematica, Tableau, Excel, SQL, and many others) come down to three
inter-related factors:
  * It is immensely powerful.
  * It is open-source.
  * It is very and increasingly widely used.

# R: A power tool for data analysis

The range and depth of statistical analyses and general data analyses that can
be accomplished with R is immense.

* Built into R are virtually the entire repertoire
  of widely known and used statistical methods. 

* Also built in to R is an extensive graphics library.

* R has a vast set of add-on or contributed
  packages. As of `r format(Sys.Date(), "%B %Y")`, there are 
  `r nrow(available_packages)` additional contributed
  packages.  
  
* R is a programming language that is specialized to efficiently
  manipulate and perform calculations on data.  

* The R programming language itself can be extended by interfacing with other
  programming languages like C, C++, Fortran, Python, and high performance computing or big
  data tools like Hadoop, Spark, SQL. 

# R: Open source software

* R is free and open source software, distributed according to the GNU
public license. 
* Likewise, virtually all of contributed R packages are likewise free and open source.
* In practical terms, this means that is freely available for everyone to use, now and forever, on more
or less any device they choose.
* Open source software always has the potential to *go viral* and
develop a large self-sustaining community of user/developers. This has arguably happened with R.

# R: Popularity and widespread use

* When it comes to the computational implementation of modern statistical
methods, R is the de facto standard. For example, the
\href{https://www.jstatsoft.org}{Journal of Statistical Software} is
overwhelmingly dominated by programs written in R.
* R is also currently very highly ranked according to many rankings of
widely used programming languages of any kind. It ranked in the top 10 or top 20 
most widely used programming languages. 
* R is ranked as one of the top five
most popular data science programs in jobs for data scientists, and in
multiple surveys of data scientists, it is often ranked as the first or
second mostly widely used data science tool.

# Introducing R commands
```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, warning = FALSE, message = FALSE, comment='#>')

# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)

```

* A useful way to think about R, and not an inaccurate one either, is that it is
simply a calculator.
```{r}
2 + 2 # addition
3 - 5 # subtraction
3 * 2 # multiplication
4 / 3 # division
(2 + 2) ^ (3 / 3.5) # exponents and brackets
```

#  Equality/inequality operations

* Testing for the equality or inequality of pairs of numbers,
already starts to go beyond the usual capabilities of handheld calculator.

```{r}
12 == (6 * 2)         # test for equality
(3 * 4) != (18 - 7)   # test for inequality
3 < 10                # less than
(2 * 5) <= 10         # less than or equal
```

# Logical values and logical operations

* In the previous step, the results are returned as either `TRUE` or `FALSE`. These are logical or *Boolean*
values.
* Just as we can represent numbers and operations on
numbers, so too can we have two logical values, `TRUE` and `FALSE` (always written in
all uppercase), and Boolean operations (*and*, *or*, and *not*) on logical values.
```{r}
TRUE & FALSE  # logical and
TRUE | TRUE   # logical or
!TRUE         # logical not
(TRUE | !TRUE) & !FALSE
```

# Variables and assignment

* If we type the following at the command prompt and then press Enter, the result is displayed but not stored.
```{r}
(12/3.5)^2 + (1/2.5)^3 + (1 + 2 + 3)^0.33
```

* We can, however, assign the value of the above calculation to a variable named `x`.
```{r}
x <- (12/3.5)^2 + (1/2.5)^3 + (1 + 2 + 3)^0.33           
```

* Now, we can use `x` as is it were a number.
```{r}
x
x ^ 2
x * 3.6
```

# Assignment rules

* In general, the assignment rule is 
```
name <- expression
```
The `expression` is any R code that returns some value.

* The `name` must consist of letters, numbers, dots, and underscores. 
``` 
x123   # acceptable
.x
x_y_z
xXx_123
```
* It must begin with a letter or a dot that is not followed by a number. 
``` 
_x   # not acceptable
.2x 
x-y-z
```
* The recommendation is to use names that are
meaningful, relatively short, without dots (using `_` instead for punctuation), and primarily consisting 
of lowercase characters.


# Vectors

* Vectors are one dimensional sequences of values.
* For example, if we want to
create a vector of the first 6 primes numbers, we could do the following.
```{r}
primes <- c(2, 3, 5, 7, 11, 13)
```
* We can now perform operations (arithmetic, logical, etc) on the `primes` vector.
```{r}
primes + 1
primes / 2
primes == 3
primes >= 7
```

# Indexing vectors

* For any vector, we can refer to individual elements using indexing
operations.  
```{r}
primes[1]
primes[5]
```

* If we want to refer to sets of elements, rather than just individual elements,
we can use vectors (made with the `c()` function) inside the indexing square
brackets. 
```{r}
primes[c(3, 5, 2)]
```

* If we use a negative valued index, we can refer to or all elements *except* one.
```{r}
primes[-1]
primes[-2]
```

# Vector types

* A vector be a sequence of numbers, logical values, or characters.
```{r}
nation <- c('ireland', 'england', 'scotland', 'wales')
```
* We can index this vector as we did with a vector of numbers.
```{r}
nation[1]
nation[2:3]
nation == 'ireland'
```

* The `class` function in R will identify the data type of the vector.
```{r}
class(primes)
class(nation)
class(nation == 'ireland')
```

# Data frames

* Data frames are rectangular data structures; they have certain number of
columns, and each column has the same number of rows. Each column is in fact a
vector.

* Usually, data frames are created when read in the contents of a data file, but we can
produce them on the command line with the `data.frame()`.
```{r}
Df <- data.frame(name = c('billy', 'joe', 'bob'), 
                 age = c(21, 29, 23))
Df
```

# Indexing data frames

* We can refer to elements of a data frame in different ways.
* The simplest is to use double indices, one for the rows, one for the columns. 
```{r}
Df[3, 2] # row 3, col 2
Df[c(1, 3), 2] # rows 1 and 3, col 2
Df[1,] # row 1, all cols
Df[, 2] # all rows, col 2
```


# Indexing data frames (contined)

* We could also refer to the column by name. To do so, we
could use the following `$` notation.
```{r}
Df$age
```
* An alternative syntax that accompishes the same thing is to use
*double* square brackets as follows.
```{r}
Df[['age']]
```
* A *single* square brackets, we would
obtain the following.
```{r}
Df['age']
```

# Functions

* In functions, we put data in, calculations or
done to or using this data, and new data, perhaps just a single
value, is then returned.
* There are probably hundreds of thousands of functions in R.
* For example, 
```{r}
length(primes)
sum(primes)
mean(primes)
median(primes)
sd(primes)
var(primes)
```

# Custom functions

* R makes it easy to create new functions. 
```{r}
my_mean <- function(x){ sum(x)/length(x)}
```
* This `my_mean` takes a vector as
input and divides its sum by the number of elements in it. It then returns this values. The `x` is a
placeholder for whatever variable we input into the function. 
* We would use it just as we would use `mean`.
```{r}
my_mean(primes)
```


# Writing R scripts

* Scripts are files where we write R commands, which can be then saved for later
use.
* You can bring up RStudio's script editor with Ctrl+Shift+N, or go to the
File/ New File/ R script, or click on the New icon on the left of
the taskbar below the menu and choose R script.

* In a script, you can have as many lines of code as you wish, and there can be
as many blank lines as you wish.  
```{#numCode .R .numberLines}
composites <- c(4, 6, 8, 9, 10, 12)

composites_plus_one <- composites + 1

composites_minus_one <- composites - 1
```
* If you place the cursor on line 1, you can then click the Run icon, or press the Ctrl+Enter keys.

# Writing R scripts (continued)

One reason why writings in scripts is very practically
valuable, even if you don't wish to save the scripts, is when you are write long
and complex commands.  
```{.R .numberLines}
Df <- data.frame(name = c('jane', 'joe', 'billy'),
                 age = c(23, 27, 24),
                 sex = c('female', 'male', 'male'),
                 occupation = c('tinker', 'tailor', 'spy')
)
```
We can execute this command as if it were on a single line by placing the cursor
anywhere on any line and pressing Ctrl+Enter.

# Code comments

* An almost universal feature of programming language is the option to
write *comments* in the code files. 
* A comment allows you write to notes or
comments around the code that is then skipped over when the script or the code
lines are being executed. 
* In R, anything following the `#` symbol on any line is 
treated as a comment. 
```{.R .numberLines}
# Here is a data frame with four variables.
# The variables are name, age, sex, occupation.
Df <- data.frame(name = c('jane', 'joe', 'billy'),
                 # This line is a comment too.
                 age = c(23, 27, 24), # Another comment. 
                 sex = c('female','male', 'male'),
                 occupation = c('tinker', 'tailor', 'spy')
)

```

# Packages

* There are presently 
  `r nrow(available_packages)` contributed
  packages in R.
  
* The easiest way to install a package is to click the  Install button on
the top left of the *Packages* window in the lower right pane.

* You can also install a package or packages with the `install.packages` command.
```{r, eval=FALSE}
install.packages("dplyr")
install.packages(c("dplyr", "tidyr", "ggplot2"))
```
  
* Having installed a package, it must be loaded to be used. This can be done by clicking the tick box before the package name in the *Packages* window, or use the `library` command.
```{r, eval=FALSE}
library("tidyverse")
```
  
# Reading in data

* R allows you to import data from a very large
variety of data file types, including from other statistics programs like SPSS,
Stata, SAS, Minitab, and so on, and common file formats like `.xlsx` and `.csv`.

* When learning R initially, the easiest way to import data is using the Import Dataset button in the Environment window.

* If we use the *From Text (readr)...* option, it runs the `read_csv` R command, which we can run ourselves on the command line, or write in a script.

```{r}
library(readr)
test_data <- read_csv("data/weight.csv")
```

# Viewing data


* The easiest way to view a data frames is to type its name.


```{r}
test_data
```


# Viewing data (continued)

* Another option to view a data frame is to `glimpse` it.

```{r}
library(dplyr)
glimpse(test_data)
```

